---
title: "Module 2 Exercises: Introduction to R and RStudio"
format: html
editor: visual
---

In this module you will get oriented inside RStudio, import `chem_data.csv`, and work through a structured data validation checklist. The emphasis is on understanding *what is happening* at each step, not just running code and moving on.

Three check-ins are interspersed with the lecture: **Check-in 2A** follows section 2.1 (Why Use a Programmatic Tool?), **Check-in 2B** follows section 2.3 (Importing Data with Explicit Column Types), and **Check-in 2C** follows section 2.4 (Data Validation Checklist). Do not work ahead.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
chem <- read_csv("../../data/chem_data.csv")
```

---

## Check-in 2A: Getting Oriented in RStudio *(8 minutes — after section 2.1)*

### Setup Steps

Before answering any questions, do the following in order:

1. Open the course `.Rproj` file from your file system (do not just open the `.qmd` file directly).
2. In the **Console** pane, type `1 + 1` and press Enter. Confirm you see `[1] 2`.
3. Return to this document in the **Source** pane. Run the setup chunk at the top of this file (the one that loads `tidyverse` and imports `chem`). Use the green Run button or Ctrl/Cmd+Shift+Enter.

### RStudio Panes

Fill in the table below. The first column is provided. Leave the last two columns blank and fill them in based on what you observe on your screen.

| Pane name | What it contains | One specific thing you notice right now |
|---|---|---|
| Source | | |
| Console | | |
| Environment | | |
| Files / Plots / Help | | |

### Written Questions

**Question 1.** After running the setup chunk, what appears in the Environment pane? Be specific about what information is shown.

> Your answer:

**Question 2.** Why does opening the `.Rproj` file first matter, rather than just opening the `.qmd` file directly?

> Your answer:

**Question 3.** You run a line of code in the Console that produces a result. Is that result saved anywhere you can use later? How is this different from running the same line inside a code chunk in this document?

> Your answer:

---

**Stop here** — your instructor will debrief before continuing to section 2.2.

---

## Check-in 2B: Importing and Inspecting chem_data *(20 minutes — after section 2.3)*

### Part 1 — Run and Interpret

Run each of the three chunks below in order. After each one, answer the question that follows in the space provided.

```{r}
#| label: glimpse-chem
glimpse(chem)
```

**How many rows does `chem` have? How many columns? Which columns currently have type `<chr>` that you would expect to be stored as something else?**

> Your answer:

---

```{r}
#| label: head-chem
head(chem, 10)
```

**What does each row in this dataset represent? Write one precise sentence.**

> Your answer:

---

```{r}
#| label: summary-chem
summary(chem)
```

**What is unusual about `pregnancy_status_at_exam` in this summary? What does the summary of `value` tell you about the range of results in this dataset?**

> Your answer:

---

### Part 2 — Re-import with Explicit Column Types

When you use `read_csv()` without specifying column types, R makes educated guesses. Those guesses are often wrong — or at least imprecise. The code below re-imports `chem_data.csv` with explicit types.

Fill in the two blanks marked with `___`. One is a format string for `col_datetime()`; the other is the name of the `col_` function appropriate for `pat_type`. Hint: look at what the other factor columns are using.

```{r}
#| label: reimport-typed
chem <- read_csv("../../data/chem_data.csv",
  col_types = cols(
    gender                   = col_factor(),
    pregnancy_status_at_exam = col_factor(),
    collect_dt               = col_datetime("___"),    # fill in the format string
    receive_dt               = col_datetime("%Y/%m/%d %H:%M:%S"),
    verify_dt                = col_datetime("%Y/%m/%d %H:%M:%S"),
    pat_type                 = ___(),                  # fill in the col_ function
    payor_groups             = col_factor(),
    test                     = col_factor(),
    units                    = col_factor()
  ))
glimpse(chem)
```

**Which columns changed type compared to the first import? What did they change from and to?**

> Your answer:

---

**Stop here** — your instructor will debrief before continuing to section 2.4.

---

## Check-in 2C: Data Validation Checklist *(15 minutes — after section 2.4)*

Before analyzing any dataset, you should work through a short validation checklist and record what you find. Think of this as a signed receipt: you are documenting that the data arrived in the shape you expected.

Run each chunk below. After each one, write your interpretation in the space provided — not just the number, but what it *means* given what you know about this dataset.

---

```{r}
#| label: val-rowcount
# 1. Row count — does it match what we expect?
nrow(chem)
```

*Your interpretation:*

---

```{r}
#| label: val-patients
# 2. Distinct patients — how many unique individuals?
n_distinct(chem$mrn)
```

*Your interpretation:*

---

```{r}
#| label: val-analytes
# 3. Analyte coverage — are all 24 analytes present and evenly represented?
count(chem, test) |> print(n = 30)
```

*Your interpretation:*

---

```{r}
#| label: val-pattype
# 4. Patient type — what categories exist and how common is missing data?
count(chem, pat_type, sort = TRUE)
```

*Your interpretation:*

---

```{r}
#| label: val-values
# 5. Value range — any implausible results?
chem |>
  group_by(test) |>
  summarize(
    min       = min(value, na.rm = TRUE),
    max       = max(value, na.rm = TRUE),
    n_missing = sum(is.na(value))
  ) |>
  print(n = 30)
```

*Your interpretation:*

---

```{r}
#| label: val-timestamps
# 6. Timestamp completeness
chem |> summarize(
  missing_collect = sum(is.na(collect_dt)),
  missing_receive = sum(is.na(receive_dt)),
  missing_verify  = sum(is.na(verify_dt))
)
```

*Your interpretation:*

---

### Closing Question

Based on your validation work above, identify **one data quality issue** you would want to investigate further before presenting results to a lab director. Describe the issue and explain why it matters.

> Your answer:

---
