---
title: "Module 03 - Visualization in R"
format: html
editor: visual
---

## Setup Code Chunk

Run this to reload your *chem* dataframe and set up the additional data sets we will use in the exercises.

```{r setup}
## remove all user defined objects in your workspace
rm(list=ls())

library(tidyverse)  # Provides functions used throughout this session

chem <- read_csv("../data/chem_data.csv", 
    col_types = cols(gender = col_factor(),
                     pregnancy_status_at_exam = col_factor(),
                     collect_dt = col_datetime("%Y/%m/%d %H:%M:%S"), 
                     receive_dt = col_datetime("%Y/%m/%d %H:%M:%S"), 
                     verify_dt = col_datetime("%Y/%m/%d %H:%M:%S"), 
                     pat_type = col_factor(),
                     payor_groups = col_factor(),
                     test = col_factor(),
                     units = col_factor())) # Read in data - note we determined data types up front
# prepare a data set that contains only potassium values
k_data <- chem |> 
  filter(test == 'potassium')
# prepare a data set that contains electrolytes commonly assayed in a metabolic panel
lyte_data <- chem |> 
  filter(test %in% c('sodium', 'potassium', 'chloride', 'bicarbonate'))
# prepare a wide version of the data set
chem_wide <- chem |> 
  pivot_wider(id_cols = c(mrn, gender, age, pregnancy_status_at_exam, pat_type, payor_groups,
                          collect_dt, receive_dt, verify_dt),
              names_from = test,
              values_from = value)
```

**Stop Here**

## Creating a ggplot

```{r}
ggplot(data = k_data) + 
  geom_histogram(mapping = aes(x = value))
```

Does the plot fit the image you were expecting? Why or why not?

**Stop Here**

## The ggplot Template and Aesthetics

Let's recreate the histogram of potassium results.

You will need to:

1.  Pick a data set: *k_data*

2.  Choose a geom function: `geom_histogram()`

3.  Map aesthetic properties of the graph to variables of the data frame (in this case, the *value* variable should be on the x axis).

Fill in the blanks and run the code to create the plot.

(Hint: if you get stuck, take a look at the slide titled "ggplot()"!)

```{r}
#k_data <- chem |> # uncomment these lines if you do not have k_data in your environment
#  filter(test == 'potassium')
ggplot(data = ______) +
  ______(mapping = aes(x = ______))
```

When you run this code, you will most likely have what looks like an error displayed above the graph:

\`stat_bin()\` using \`bins = 30\`. Pick better value with \`binwidth\`.

This is actually just a message from R, telling you that it's a bad idea to use a default bin width.

The *value* variable holds the quantitative value a given test in this data set. For this specific analyte, what would be an appropriate binwidth for *value* to see a distribution aligned to the likely resolution of the measurements? (Visually inspect the data if you are less familiar with the range of typical potassium values.)

Now fill in the blanks below to recreate the code from the previous chunk, with the *binwidth* argument passed to the *geom_histogram* function.

```{r}
ggplot(______) +
  ______(mapping = aes(______), binwidth = ______)
```

The effect of reducing the bin width (and increasing the number of bins) is that the data is now presented in much greater detail.

Let's next add some color!

Create a new code chunk below. Copy and paste the code from the previous code chunk and add an aesthetic mapping that maps the *fill* aesthetic to the *gender* variable. (Hint: all aesthetics must be defined \*inside\* the aes() function!)

**Stop Here**

## Geom Functions

Run the following code chunk:

```{r}
ggplot(data = k_data) +
  geom_histogram(mapping = aes(x = value), binwidth = 0.1)
```

Now try to figure out how you would modify the code above so it draws a frequency polygon instead of histogram. One resource is the ggplot cheatsheet that can be found at <https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf>

```{r}

```

Modify the previous code chunk so that the frequency polygon's line color is blue. (Hint: color names go in quotes!)

```{r}

```

What do you think the following code will do? Try to predict what you'll see.

Then run it.

```{r}
ggplot(data = k_data) +
  geom_histogram(mapping = aes(x = value), binwidth = 0.1) +
  geom_freqpoly(mapping = aes(x = value), binwidth = 0.1)
```

**Stop Here**

## Small Multiples for Comparisons

A grid of subplots can be produced using faceting functions, including `facet_grid()` which we demonstrate below. You can think about faceting like grouping the data frame - it is a convenient method to generate plots of subsets of data. The code chunk below takes our *k_data* dataframe and generates a series of histograms that are split into 6 columns based on the *pat_type*.

```{r}
ggplot(data = k_data) +
  geom_histogram(mapping = aes(x = value), binwidth = 0.1) +
  facet_grid(cols = vars(pat_type),
             scales = "free")
```

**What other variables could be used to split the data set further? Fill in the code chunk below to create a grid of plots for the potassium values based on two qualitative variables.**

```{r}
ggplot(data = k_data) +
  geom_histogram(mapping = aes(x = value), binwidth = 0.1) +
  facet_grid(rows = vars(______),
             cols = vars(______),
             scales = "free")
```

**Stop Here**

## Quantitative Comparisons

In the setup chunk we created a wide version of the data to help compare quantitative values (if you want to learn more about how copy and paste code into Claude and ask for an explanation). Review the first few rows of data below.

```{r}
head(chem_wide)
```

We now have a data set where every lab value for a single patient is contained on one row. This can help perform visual comparisons between values, as each variable (column) can be provided to ggplot as input to the plot. We can use `geom_point()` to build a scatterplot with ggplot and compare ALT with AST, two liver enzymes that would be expected to correlate with one another.

```{r}
ggplot(data = chem_wide) +
  geom_point(mapping = aes(x = alanine_aminotransferase, y = aspartate_aminotransferase))
```

**What other variables (tests) may have interesting relationships in this data set? Fill in the code chunk below to visualize the relationship between 2 other tests.**

```{r}
ggplot(data = chem_wide) +
  geom_point(mapping = aes(x = ______, y = ______))
```

**Are there other variables that may be interesting to visualize? How would you incorporate them into the visualization to visualize 3 variables at once? Use the code chunk below to add another variable.** The RStudio ggplot Cheatsheet may be helpful: <https://opensource.posit.co/resources/cheatsheets/data-visualization/>. Consult Claude for help if you have an idea but are not sure what the correct syntax is, and be sure to include prior code as well as additional information not represented in the code (e.g. variable names).

```{r}

```

## Let's Put It To Practice

**Using Claude for assistance, optimize the faceted plot you created in the Small Multiples exercise for readability and interpret-ability.** Consider the overall color scheme, background, labels, and titles, in addition to any other relationships that may be of interest to visualize.

```{r}
ggplot(data = k_data) +
  geom_histogram(mapping = aes(x = value), binwidth = 0.1) +
  facet_grid(rows = vars(______),
             cols = vars(______),
             scales = "free")
```

**Now optimize the scatterplot you developed, considering similar principles as above.**

```{r}
ggplot(data = chem_wide) +
  geom_point(mapping = aes(x = ______, y = ______))
```
