---
title: "Module 5 Exercises: Computational Thinking and Reproducibility [SUPERSEDED]"
format: html
editor: visual
---

> **Note:** In the updated 4-module course structure (Option 2), the content from this file has been reorganized:
>
> - The CT decomposition exercise (Exercise 5A) → **Check-in 4A** in `module_4_exercises.qmd`
> - The reproducible mini-report (Exercise 5B) → **Exercise 4H** in `module_4_exercises.qmd` (expanded and reworked to use the LC-MS/MS dataset)
>
> This file is retained for reference only. Use `module_4_exercises.qmd` for the current course.

---

This final module asks you to zoom out. Rather than filling in individual blanks, you will decompose a real lab problem into its logical steps, build a small reproducible report, and reflect on what it means for an analysis to be reproducible.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
chem <- read_csv("../../data/chem_data.csv",
  col_types = cols(
    gender     = col_factor(),
    collect_dt = col_datetime("%Y/%m/%d %H:%M:%S"),
    receive_dt = col_datetime("%Y/%m/%d %H:%M:%S"),
    verify_dt  = col_datetime("%Y/%m/%d %H:%M:%S"),
    pat_type   = col_factor(),
    test       = col_factor()
  ))
```

---

## Exercise 5A: Decompose a Lab Problem

*Estimated time: 15 minutes*

> **Scenario:** The lab director asks: *"Which analytes had a turnaround time over 4 hours for more than 5% of results in this dataset, broken down by patient type?"*

Before writing any code, work through the four thinking steps below. This is the reasoning process that a good data analyst does before touching a keyboard.

---

### Decompose

Break the lab director's question into smaller sub-questions that each have a concrete answer. A good decomposition makes each step feel manageable and reveals hidden assumptions.

What sub-questions do you need to answer in sequence to solve the full problem?

> 1.
> 2.
> 3.
> 4.
> 5.

---

### Pattern Recognition

What features of `chem_data`'s structure make this problem solvable? In other words, what information does the dataset already contain that you will rely on?

> Your answer:

---

### Abstraction

In a solution to this problem, what stays *constant* regardless of the specific question being asked, and what would you change to answer a related but different question (for example, "more than 2 hours" or "more than 10%" or "broken down by gender instead of patient type")?

> What stays constant (the structure of the pipeline):

> What varies (the parameters you would swap out):

---

### Algorithm

Write plain-English pseudocode for a solution — step by step. Do not use R syntax. Write it the way you would explain the process to a colleague who has never used R.

```
Step 1:
Step 2:
Step 3:
Step 4:
Step 5:
Step 6:
```

---

### Working Solution

The chunk below contains R code that implements exactly what the lab director asked for. Run it and compare each line to your pseudocode.

```{r}
#| label: tat-solution
#| eval: false
# Compare this to your pseudocode — does each step correspond to a line?
chem |>
  mutate(tat_hours = as.numeric(verify_dt - collect_dt, units = "hours")) |>
  filter(!is.na(pat_type)) |>
  group_by(test, pat_type) |>
  summarize(
    n_total    = n(),
    n_long_tat = sum(tat_hours > 4, na.rm = TRUE),
    pct_long   = round((n_long_tat / n_total) * 100, 1),
    .groups    = "drop"
  ) |>
  filter(pct_long > 5) |>
  arrange(desc(pct_long))
```

**Change `eval: false` to `eval: true` in the chunk options above (or copy it into the Console to run it). Then answer:**

**Are any of the analyte/patient-type combinations in the results surprising? What might explain why inpatient or ICU samples have a higher percentage of long TAT results compared to outpatient samples?**

> Your answer:

---

**Stop Here** — your instructor will pause to debrief the group before continuing.

---

## Exercise 5B: Build a Reproducible Mini-Report

*Estimated time: 30 minutes*

This exercise is a structured Quarto authoring experience. Each section below has a code chunk for you to complete and a space for a brief narrative paragraph. After completing all four sections, render the document to HTML using the Render button at the top of RStudio.

Then: change the analyte name in Section 2 from `"potassium"` to any other analyte in the dataset, and re-render the document. Observe what updates automatically.

---

### Section 1 — Data Validation

Fill in the two blanks, then run the chunk and write your interpretation.

```{r}
#| label: s1-validation
#| message: false
# Run your validation checklist — fill in the blanks
nrow(chem)
n_distinct(chem$___)
count(chem, test) |> filter(n != ___)   # flag any analytes with unexpected row counts
```

*Narrative: Write 2–3 sentences describing what the validation showed and whether you have any concerns.*

> Your narrative:

---

### Section 2 — Analyte Selection

The variable `selected_analyte` controls the rest of the report. For now, leave it set to `"potassium"`. After you have completed and rendered the full document once, come back here to change it.

```{r}
#| label: s2-analyte-selection
# Choose an analyte — change this value to explore a different one
selected_analyte <- "potassium"   # <-- change me

analyte_data <- chem |>
  filter(test == ___)
```

*Narrative: Briefly describe the analyte you chose and any clinical context that is relevant — what does it measure, what are typical clinical concerns?*

> Your narrative:

---

### Section 3 — Summary by Patient Type

Fill in the four blanks to compute a summary table grouped by patient type.

```{r}
#| label: s3-summary
analyte_summary <- analyte_data |>
  filter(!is.na(___)) |>
  group_by(___) |>
  summarize(
    n            = n(),
    mean_value   = mean(value,   na.rm = TRUE),
    median_value = ___,
    sd_value     = sd(value, na.rm = TRUE)
  )
analyte_summary
```

*Narrative: What does the summary tell you? Are there meaningful differences across patient types in either the typical value or the spread? What clinical explanation might there be?*

> Your narrative:

---

### Section 4 — Distribution

Fill in the three blanks to create a faceted histogram. The `binwidth` should be appropriate for your selected analyte's measurement scale — try a few values and choose one that shows the shape clearly without being too noisy or too coarse.

```{r}
#| label: s4-distribution
ggplot(analyte_data |> filter(!is.na(pat_type)),
       aes(x = value, fill = pat_type)) +
  geom_histogram(binwidth = ___, alpha = 0.6, position = "identity") +
  facet_wrap(~ pat_type, ncol = 2) +
  theme_bw() +
  labs(title = paste(___, "results by patient type"),
       x     = paste0(___, " (", unique(analyte_data$units), ")"),
       y     = "Count")
```

*Narrative: Describe the distribution. Is it roughly symmetric or skewed? Are there outliers visible in any patient type group? Does the shape differ meaningfully across groups?*

> Your narrative:

---

### Closing Reflection

After re-rendering with a different analyte, answer these questions:

**What updated automatically when you changed `selected_analyte`?**

> Your answer:

**What did you have to update manually?**

> Your answer:

**What would need to change to make this report fully *parameterized* — meaning the analyte name appears in only one place in the entire document, and changing it there updates everything, including axis labels, titles, and narrative text?**

> Your answer:

---
