---
title: "Module 4 Exercises: Computational Thinking"
format: pdf
editor: visual
execute:
  eval: false
---

# Computational Thinking with GenAI:

# LEARNER WORKBOOK

For this lesson, all the code you need is here or comes from Claude.

### How to use this file:

\- Run ONE section at a time. Highlight the lines, then press Ctrl+Enter (Windows) or Cmd+Enter (Mac).

\- Paste code from Claude where it says PASTE HERE.

\- If code gives an error, copy the full error message to Claude.

\- Put chem_data.csv in the same folder as this file, then use

Session \> Set Working Directory \> To Source File Location.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
chem <- read_csv("../../data/chem_data.csv")
```

### Scenario:

### ICU clinicians say their chemistry results come back slower

As the lab director, you've requested data to help you determine

Q1. Does turnaround time (TAT) differ by patient type?

Q2. Are we meeting a 3-hour collect-to-verify goal?

# 

```{r}
# RUN THIS CODE CHUNK TO LOAD REQUIRED PACKAGES
library(tidyverse)

```

## Build your data subset

Fill in the two blanks with "potassium" and "creatinine". Keep the quote marks. This will create a data set that only includes data from glucose, potassium, and creatinine tests.

```{r}
tests_keep <- c("glucose", , )

read_csv("chem_data.csv", show_col_types = FALSE) |> 
    filter(test %in% tests_keep) |> 
    write_csv("chem_subset.csv")

chem <- read_csv("chem_subset.csv", show_col_types = FALSE) 

glimpse(chem)

```

CHECK: you should have 19,443 rows (6,481 patients x 3 tests).

# 

## INITIAL GEN AI PROMPT

Do not upload the data file to Claude.

Paste your prompt here:

Paste the code Claude gave you in the code chunk below. Run it. What happened?

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

# 

## EXERCISE 1. Decomposition (Objective 1)

No code. Follow the slides. You can type your response below.

## EXERCISE 2. Pattern recognition (Objective 2)

The code below works for ONE test (glucose) in ONE patient type (icu). Run it first.

```{r}
chem |> filter(test == "glucose", pat_type == "icu") |> 
          summarize(n_results = n(), median_value = median(value))

```

### A) Claude's version for every patient type, glucose only

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

### B) Claude's version for every test and every patient type

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

## EXERCISE 3. Abstraction (Objective 3)

### Run both code chunks below. Then inspect the code in each chunk to find what differs between them.

```{r}
chem |> filter(test == "glucose") |> 
          group_by(pat_type) |> 
          summarize(n_results = n(), pct_above = round(100 * mean(value > 125), 1))


```

Answer:

```{r}
chem |> filter(test == "potassium") |> 
          group_by(pat_type) |> 
          summarize(n_results = n(), pct_above = round(100 * mean(value > 5.0), 1))


```

Answer:

What differs between the two code chunks?

### Claude's code from your filled-in prompt template

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

# 

## EXERCISE 4. Algorithm design (Objective 4)

## REVISED GEN AI PROMPT

Paste your prompt here:

### Paste the code Claude wrote AFTER you answered its questions and approved its plan.

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

CHECK: the patient counts in your table should total 6,481, and patients with no patient type should still be there as a group.

# 

### DEBRIEF. Does this make sense?

```{r}
n_distinct(chem$mrn) # how many patients are in the data? Do they total 6,481?

nrow(chem)/n_distinct(chem$mrn)    # how many rows per patient? Is there 3 (one for each test)


```

If the total number of patients is 5,393, patients with no patient type were dropped.
