---
title: "Module 3 Exercises: Visualization Principles and Best Practices"
format: html
editor: visual
---

This module builds on the import and validation skills from Module 2. You will build charts, identify flaws in poorly designed visualizations, and practice distinguishing exploratory from explanatory work.

The setup chunk below creates three filtered subsets for the coding exercises. It uses `filter()` — a function we will learn fully in Module 4. For now, treat these lines as a black box that hands you the right data for each exercise.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)

chem <- read_csv("../../data/chem_data.csv",
  col_types = cols(
    gender     = col_factor(),
    collect_dt = col_datetime("%Y/%m/%d %H:%M:%S"),
    verify_dt  = col_datetime("%Y/%m/%d %H:%M:%S"),
    pat_type   = col_factor(),
    test       = col_factor()
  ))

# We'll understand filter() in Module 4 — for now these just create
# the subsets you need for each exercise.
k_data      <- chem |> filter(test == "potassium")
lyte_data   <- chem |> filter(test %in% c("sodium", "potassium", "chloride", "bicarbonate"))
sodium_data <- chem |> filter(test == "sodium")
```

---

## Check-in 3A: Exploratory or Explanatory? *(5 minutes — after section 3.2)*

Your instructor will display three charts on screen. For each one, classify it as **exploratory** or **explanatory** and write one sentence explaining what signal in the chart led to your decision.

> **Chart 1:**
>
> Classification:
>
> Reason:

> **Chart 2:**
>
> Classification:
>
> Reason:

> **Chart 3:**
>
> Classification:
>
> Reason:

**Reflection:** Is it possible for the same chart to be used exploratorily by one person and explanatorily by another? Give an example from the MS lab context.

> Your answer:

---

## Check-in 3B: Match the Chart *(5 minutes — after section 3.3)*

For each scenario below, write the chart type you would use and one sentence of justification. There is often more than one defensible answer — focus on the reasoning.

> **Scenario 1:** You want to see how sodium results are distributed across the ICU, inpatient, and outpatient groups simultaneously.
>
> Chart type:
>
> Why:

> **Scenario 2:** You are reviewing one month of daily QC results for a potassium assay and want to spot runs or trends.
>
> Chart type:
>
> Why:

> **Scenario 3:** You want to show a poster audience how measured concentration relates to expected concentration across your calibration curve.
>
> Chart type:
>
> Why:

> **Scenario 4:** You want to quickly check whether the 24 analytes in chem_data have plausible result distributions before proceeding with any analysis.
>
> Chart type:
>
> Why:

> **Scenario 5:** You need to compare the median TAT across five patient type groups in a written report to lab leadership.
>
> Chart type:
>
> Why (and is a table ever the better choice here?):

---

## Check-in 3C: Spot the Flaws *(8 minutes — after section 3.4)*

Your instructor will display two charts on screen. **Do not run any code yet** — just examine what you see and answer the questions below. You will fix both charts in Exercise 3E.

> **Bad Chart 1** (a bar chart of analyte means):
>
> Flaw 1 — what is wrong and what does it cause a viewer to misunderstand?
>
> Flaw 2:
>
> Flaw 3:

> **Bad Chart 2** (a boxplot of sodium results by patient type):
>
> Flaw 1 — what is wrong and what does it cause a viewer to misunderstand?
>
> Flaw 2:
>
> Flaw 3:

---

## Exercise 3D: First Plots with ggplot2 *(20 minutes — after section 3.5)*

*This is the first coding exercise in the module. Work through each step in sequence; your instructor will circulate.*

### Step 1 — Skeleton Histogram

Fill in the two blanks to create a histogram of potassium results. `x` should map to the result value; `binwidth` controls bar width.

```{r}
#| label: hist-potassium-basic
ggplot(k_data, aes(x = ___)) +
  geom_histogram(binwidth = ___)
```

Try `binwidth = 0.1`, `0.5`, and `1.0`. Which best represents the data given that potassium is reported to one decimal place? Why?

> Your answer:

---

### Step 2 — Color by Patient Type

Fill in the three blanks: the `fill` aesthetic, a descriptive `title`, and a `y`-axis label.

```{r}
#| label: hist-potassium-pattype
ggplot(k_data, aes(x = value, fill = ___)) +
  geom_histogram(binwidth = 0.1, alpha = 0.6, position = "identity") +
  theme_bw() +
  labs(title = "___",
       x     = "Potassium (mEq/L)",
       y     = "___",
       fill  = "Patient Type")
```

What do you notice about the distribution of potassium across patient types? Does anything look clinically plausible — for example, do ICU patients differ from outpatients in a way that makes sense?

> Your answer:

---

### Step 3 — Exploratory or Explanatory?

Is the chart you just made an *exploratory* visualization or an *explanatory* one? What would you need to change to make it suitable for a conference poster or a quality report?

> Your answer:

---

**Stop here** — your instructor will pause to debrief before continuing.

---

## Exercise 3E: Distributions, Run Chart, and Bad Chart Rewrites *(22 minutes — after Exercise 3D debrief)*

### Part 1 — Comparing Distributions and a Run Chart *(8 minutes)*

**Boxplots with faceting.** Fill in the two blanks to create side-by-side boxplots for the four electrolytes, with patient type on the x-axis and one panel per analyte.

```{r}
#| label: boxplot-lytes
ggplot(lyte_data |> filter(!is.na(pat_type)),
       aes(x = ___, y = value)) +
  geom_boxplot() +
  facet_wrap(~ ___, scales = "free_y") +
  theme_bw() +
  labs(title = "Electrolyte distributions by patient type",
       x = "Patient type",
       y = "Result value")
```

Which analyte shows the most variation across patient types? Does that make clinical sense?

> Your answer:

**Run chart.** Fill in the two `yintercept` blanks with the lower and upper reference limits for sodium (normal range: 135–145 mmol/L).

```{r}
#| label: run-chart-sodium
# Zoom into a 4-hour window for a clearer view
sodium_window <- sodium_data |>
  filter(collect_dt >= as.POSIXct("2022-12-26 08:00:00"),
         collect_dt <  as.POSIXct("2022-12-26 12:00:00"))

ggplot(sodium_window, aes(x = collect_dt, y = value)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = ___, linetype = "dashed", color = "red") +
  geom_hline(yintercept = ___, linetype = "dashed", color = "red") +
  theme_bw() +
  labs(title = "Sodium results — morning of 2022-12-26",
       x     = "Collection time",
       y     = "Sodium (mmol/L)")
```

What elements would be needed to turn this into a proper Levey-Jennings QC chart? List at least three.

> Your answer:

---

### Part 2 — Rewrite the Bad Charts *(14 minutes)*

Both chunks below run without error. The charts are real outputs that a researcher might produce and, unfortunately, share. Your Check-in 3C analysis prepared you for this — now fix them.

**Bad Chart 1.** Run this first to see what you are working with.

```{r}
#| label: bad-chart-1
analyte_means <- chem |>
  group_by(test) |>
  summarize(mean_val = mean(value, na.rm = TRUE))

ggplot(analyte_means, aes(x = test, y = mean_val, fill = test)) +
  geom_col() +
  scale_fill_manual(values = rainbow(24)) +
  theme(axis.text.x = element_text(angle = 90))
```

Rewrite it below. You do not need to fix every flaw in a single chart — focus on the most important ones and explain your choices.

```{r}
#| label: bad-chart-1-improved
# Your improved version here

```

> Why you made these changes:

---

**Bad Chart 2.** Run this first to see what you are working with.

```{r}
#| label: bad-chart-2
ggplot(sodium_data |> filter(!is.na(pat_type)),
       aes(x = pat_type, y = value)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(138, 142))
```

Rewrite it below.

```{r}
#| label: bad-chart-2-improved
# Your improved version here

```

> Why you made these changes:

---
