---
title: "Module 4 Exercises: Computational Approaches to Data"
format: html
editor: visual
---

This module introduces a framework for thinking before you code, then applies it to the core data transformation operations. The first half uses `chem_data`; the second half introduces the course's LC-MS/MS pain management dataset — three related tables that mirror how instrument software actually exports data.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)

chem <- read_csv("../../data/chem_data.csv",
  col_types = cols(
    gender                   = col_factor(),
    pregnancy_status_at_exam = col_factor(),
    collect_dt               = col_datetime("%Y/%m/%d %H:%M:%S"),
    receive_dt               = col_datetime("%Y/%m/%d %H:%M:%S"),
    verify_dt                = col_datetime("%Y/%m/%d %H:%M:%S"),
    pat_type                 = col_factor(),
    payor_groups             = col_factor(),
    test                     = col_factor(),
    units                    = col_factor()
  ))
```

---

# FIRST HALF

## Check-in 4A: Decompose Before You Code *(8 minutes — after section 4.1)*

> **Scenario:** The lab director asks: *"Which analytes had a turnaround time over 4 hours for more than 5% of results, broken down by patient type?"*

Work through all four thinking steps **in writing, before touching any code.** A good decomposition makes the code almost write itself.

### Decompose

Break the lab director's question into smaller sub-questions, each with a concrete answer. What do you need to figure out in sequence?

> 1.
> 2.
> 3.
> 4.
> 5.

### Pattern Recognition

What features of `chem_data`'s structure make this problem solvable? Which columns will you rely on?

> Your answer:

### Abstraction

What stays *constant* regardless of the specific threshold values? What would you change to answer "more than 2 hours" or "more than 10%" instead?

> What stays constant (the pipeline structure):
>
> What varies (the parameters):

### Algorithm

Write plain-English pseudocode — step by step, no R syntax. Write it so a colleague who has never used R could follow it.

```
Step 1:
Step 2:
Step 3:
Step 4:
Step 5:
Step 6:
```

---

## Check-in 4B: Write the Filter *(8 minutes — after section 4.2)*

Fill in the four blanks to create a subset of outpatient alkaline phosphatase results for female patients aged 50 and over. Use the exact strings that appear in the data.

```{r}
#| label: filter-alk-phos
alk_phos_f <- chem |>
  filter(test     == "___",
         pat_type == "___",
         gender   == "___",
         age      >= ___)

nrow(alk_phos_f)
```

How many rows does `alk_phos_f` contain? Does that seem reasonable given the full dataset has 155,544 rows? Show your rough reasoning.

> Your answer:

Which of the four conditions was **decomposition** (narrowing to a relevant subset) and which was more like **pattern recognition** (applying a uniform criterion to all rows that pass the earlier filters)?

> Your answer:

---

## Check-in 4C: Read the Pipeline *(5 minutes — after section 4.3)*

The chunk below is pre-written. **Before running it**, predict in writing what the output will look like: how many rows, what columns, and roughly what the values in `pct_low` will represent.

```{r}
#| label: pipeline-predict
chem |>
  filter(test == "potassium", pat_type == "inpatient") |>
  mutate(low_k = value < 3.5) |>
  group_by(gender) |>
  summarize(
    n_low   = sum(low_k,  na.rm = TRUE),
    pct_low = round(mean(low_k, na.rm = TRUE) * 100, 1)
  )
```

> Your prediction (before running):

Now run it. Which step, if any, produced output that surprised you?

> Your answer:

---

## Check-in 4D: Compute TAT *(7 minutes — after section 4.4)*

Turnaround time is the elapsed time between two timestamps. Fill in the two blanks — the start and end of the interval for *total* TAT (collection to verification).

```{r}
#| label: mutate-tat
chem_tat <- chem |>
  mutate(total_tat_hours = as.numeric(___ - ___, units = "hours"))

summary(chem_tat$total_tat_hours)
```

What is the median TAT? What is the maximum? Does the maximum seem realistic for a routine chemistry result, and what might explain it?

> Your answer:

`mutate()` corresponds to **abstraction** in the CT framework. In one sentence, explain why: what rule is being expressed once and applied uniformly?

> Your answer:

---

## Check-in 4E: Group and Reshape *(10 minutes — after section 4.5)*

### Part 1 — Two Grouping Variables

Fill in the two blanks to summarize results by both analyte and patient type.

```{r}
#| label: analyte-pattype-summary
analyte_pat_summary <- chem |>
  filter(!is.na(pat_type)) |>
  group_by(___, ___) |>
  summarize(mean_value = mean(value, na.rm = TRUE),
            n          = n(),
            .groups    = "drop")

nrow(analyte_pat_summary)
```

Before running: how many rows do you expect? (How many analytes × how many patient type categories?) Show your reasoning.

> Expected rows:
>
> Actual rows after running:

### Part 2 — Reshape to Wide

Fill in the two blanks to pivot the analyte-level summary so each analyte becomes its own column.

```{r}
#| label: pivot-wide
analyte_summary <- chem |>
  group_by(test) |>
  summarize(mean_value = mean(value, na.rm = TRUE))

analyte_wide <- analyte_summary |>
  select(test, mean_value) |>
  pivot_wider(names_from  = ___,
              values_from = ___)

analyte_wide
```

Which format — long (`analyte_summary`) or wide (`analyte_wide`) — would you pipe into a `ggplot2` call? Which would you paste into a slide? Why does the answer differ?

> Your answer:

---

**Stop here — mid-module break (10 minutes)**

---

# SECOND HALF

The second half introduces the LC-MS/MS pain management dataset. This dataset has the three-table structure that clinical MS labs encounter in practice: raw peak data from the instrument, processed sample results, and per-batch calibration and metadata.

```{r}
#| label: setup-ms
#| message: false
# Three related tables from an LC-MS/MS pain management assay
peak_data   <- read_csv("../../data/peak_data.csv")
sample_data <- read_csv("../../data/sample_data.csv")
batch_data  <- read_csv("../../data/batch_data_ts.csv")

# Quick orientation
glimpse(sample_data)
glimpse(batch_data)
```

---

## Check-in 4F: Navigate the MS Dataset *(10 minutes — after section 4.6)*

Answer all four questions in writing. No code required (yet).

**Question 1.** What column(s) would you use as the key to join `sample_data` to `batch_data`? Is `batch_name` alone sufficient? How would you verify your answer before running the join?

> Your answer:

**Question 2.** Write the R code to enrich `sample_data` with instrument name and batch pass/fail from `batch_data`. You want all rows from `sample_data` regardless of whether a match exists.

```r
# Your code here

```

**Question 3.** Write the R code to identify any `batch_name` + `compound_name` combinations in `sample_data` that have *no* matching row in `batch_data`. What would such orphaned rows indicate about data quality?

```r
# Your code here

```

**Question 4.** Suppose a data entry error caused one `batch_name` + `compound_name` combination to appear 3 times in `batch_data` instead of once. How many rows would the join produce for each matching row in `sample_data`? Why is this a problem for downstream calculations?

> Your answer:

---

## Check-in 4G: Anatomy of a Quarto Document *(5 minutes — after section 4.7)*

Open the capstone file (Exercise 4H below, or a version displayed by your instructor). Find and label each of the following elements by writing its line number or a short description of where it appears:

> **YAML header** (the block at the top that controls document metadata and output format):

> **A narrative paragraph** (plain markdown text, not inside a code chunk):

> **A code chunk** (the fenced block with `{r}`):

> **A chunk option** (a line inside a chunk starting with `#|`):

> **An in-line code reference or parameterized value** (a value that would update automatically if the underlying data changed):

---

## Exercise 4H: Capstone — An LC-MS/MS Batch Review Pipeline *(44 minutes)*

This exercise builds a complete, reproducible summary of batch quality across the pain management dataset. The document has four sections. Each section has a small number of blanks and a narrative space. Work through them in order; render to HTML when all four are complete.

The pipeline structure and `|>` scaffolding are provided with step comments. Fill in the **arguments**, not the structure.

---

### Section 1 — Import and Validate *(~8 minutes)*

```{r}
#| label: s1-validate
#| message: false

# How many rows in each table?
nrow(sample_data)
nrow(batch_data)

# How many unique batches and compounds?
n_distinct(sample_data$___)      # unique batch names in sample_data
n_distinct(sample_data$___)      # unique compound names

# Any missing pass/fail values?
sum(is.na(sample_data$sample_passed))
sum(is.na(batch_data$batch_passed))
```

*Narrative: Write 2–3 sentences describing what the validation showed. Are there any mismatches between batch counts in the two tables that you would want to investigate?*

> Your narrative:

---

### Section 2 — Explore *(~8 minutes)*

Filter `sample_data` to patient samples only. The setup chunk at the top of the second half already imported `sample_data` — use it directly here.

```{r}
#| label: s2-explore
patient_samples <- sample_data |>
  filter(sample_type == "___")          # keep only patient samples

# Exploratory boxplot of measured concentration by compound,
# panels split by whether the sample passed
ggplot(patient_samples |> filter(!is.na(sample_passed)),
       aes(x = compound_name, y = concentration)) +
  geom_boxplot() +
  facet_wrap(~ ___, ncol = 1) +         # one panel per pass/fail status
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Patient sample concentrations by compound and pass/fail status",
       x = "Compound",
       y = "Concentration")
```

*Narrative: Label this visualization explicitly as exploratory. What pattern, if any, do you see between failed samples and concentration? Is there anything surprising?*

> Your narrative:

---

### Section 3 — Join and Summarize *(~12 minutes)*

Fill in the four blanks: the two key columns for the join, the grouping variables, and the CV formula. The pipeline structure is provided.

```{r}
#| label: s3-join-summarize
batch_summary <- sample_data |>
  # Step 1: enrich sample results with batch metadata
  left_join(batch_data,
            by = c("___" = "___",        # first key column (same name in both tables)
                   "compound_name" = "compound_name")) |>
  # Step 2: keep only samples from batches that passed QC
  filter(batch_passed == TRUE,
         sample_type  == "patient") |>
  # Step 3: summarize by compound and instrument
  group_by(___, ___) |>
  summarize(
    n            = n(),
    mean_conc    = round(mean(concentration,  na.rm = TRUE), 2),
    sd_conc      = round(sd(concentration,    na.rm = TRUE), 2),
    cv_pct       = round((___ / ___) * 100,  1),   # CV = SD / mean × 100
    .groups      = "drop"
  ) |>
  arrange(compound_name, instrument_name)

batch_summary
```

*Narrative: Which compound × instrument combinations have the highest CV? Does the instrument affect precision differently across compounds? Write 2–3 sentences.*

> Your narrative:

---

### Section 4 — Communicate *(~8 minutes)*

Build a polished explanatory chart from `batch_summary`. Add a horizontal reference line at 15% CV — a common acceptability threshold for LC-MS/MS assays.

```{r}
#| label: s4-communicate
ggplot(batch_summary,
       aes(x = compound_name, y = cv_pct, color = ___)) +  # color by instrument
  geom_point(size = 3) +
  geom_hline(yintercept = ___, linetype = "dashed", color = "gray40") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "___",
       x     = "Compound",
       y     = "CV (%)",
       color = "Instrument",
       caption = "___")   # one sentence explaining what a point above the line means
```

*Narrative: What would you do next if a compound × instrument combination exceeded 15% CV? Describe in one or two sentences what action this chart would prompt.*

> Your narrative:

---

### Render and Reflect *(~8 minutes)*

Render this document to HTML using the **Render** button at the top of RStudio. Once it renders successfully, go back to Section 2 and change `sample_type == "patient"` to `sample_type == "calibrator"`, then re-render.

**What updated automatically when you changed `sample_type`?**

> Your answer:

**What did you have to update manually?**

> Your answer:

**What single change would make the `sample_type` filter fully parameterized — meaning it appears in only one place in the entire document, and changing it there updates every section?**

> Your answer:

**Final reflection:** This document — raw data files in, rendered HTML report out, with no manual steps in between — is a reproducible analysis. What is the practical value of this property for a clinical MS lab performing routine batch review?

> Your answer:

---
