---
title: "Module 4 Exercises: Computational Thinking"
format: pdf
editor: visual
execute:
  eval: false
---

# Computational Thinking with GenAI:

# LEARNER WORKBOOK

For this lesson, all the code you need is here or comes from Claude.

### How to use this file:

\- Run ONE section at a time. Highlight the lines, then press Ctrl+Enter (Windows) or Cmd+Enter (Mac).

\- Paste code from Claude where it says PASTE HERE.

\- If code gives an error, copy the full error message to Claude.

\- Put chem_data.csv in the same folder as this file, then use

Session \> Set Working Directory \> To Source File Location.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
chem <- read_csv("../data/chem_data.csv", 
                 col_types = cols(gender = col_factor(),
                                  pregnancy_status_at_exam = col_factor(),
                                  collect_dt = col_datetime("%Y/%m/%d %H:%M:%S"), 
                                  receive_dt = col_datetime("%Y/%m/%d %H:%M:%S"), 
                                  verify_dt = col_datetime("%Y/%m/%d %H:%M:%S"), 
                                  pat_type = col_factor(),
                                  payor_groups = col_factor(),
                                  test = col_factor(),
                                  units = col_factor())) 
```

### Scenario:

### ICU clinicians say their chemistry results come back slower

As the lab director, you've requested data to help you determine

Q1. Does turnaround time (TAT) differ by patient type?

Q2. Are we meeting a 3-hour collect-to-verify goal?

## Build your data subset

Fill in the two blanks with "potassium" and "creatinine". Keep the quote marks. This will create a data set that only includes data from glucose, potassium, and creatinine tests.

```{r}
tests_keep <- c("glucose", ______, _______)

chem_subset <- chem |> 
    filter(test %in% tests_keep) |> 
    write_csv("chem_subset.csv")

glimpse(chem_subset)

```

CHECK: you should have 19,443 rows (6,481 patients x 3 tests).

# 

## INITIAL GEN AI PROMPT

Do not upload the data file to Claude.

Paste your prompt here:

Paste the code Claude gave you in the code chunk below. Run it. What happened?

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

# 

## EXERCISE 1. Decomposition (Objective 1)

No code. Follow the slides. You can type your response below.

- Load data

- Verify/modify data types

- Validate - missing and duplicate data

- Calculate TAT

- Run descriptive stats for TAT by patient type

- Compare TAT against 3 hr goal

## EXERCISE 2. Pattern recognition (Objective 2)

The code below works for ONE test (glucose) in ONE patient type (icu). Run it first.

```{r}
chem_subset |> filter(test == "glucose", pat_type == "icu") |> 
          summarize(n_results = n(), median_value = median(value))

```

### A) Claude's version for every patient type, glucose only

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

### B) Claude's version for every test and every patient type

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

## EXERCISE 3. Abstraction (Objective 3)

### Run both code chunks below. Then inspect the code in each chunk to find what differs between them.

```{r}
chem_subset |> filter(test == "glucose") |> 
          group_by(pat_type) |> 
          summarize(n_results = n(), pct_above = round(100 * mean(value > 125), 1))


```

Answer:

```{r}
chem_subset |> filter(test == "potassium") |> 
          group_by(pat_type) |> 
          summarize(n_results = n(), pct_above = round(100 * mean(value > 5.0), 1))


```

Answer:

What differs between the two code chunks?

### Claude's code from your filled-in prompt template

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

# 

## EXERCISE 4. Algorithm design (Objective 4)

## ALGORITHM DESIGN PART 1 - REVISED GEN AI PROMPT

Paste your prompt here:

### Paste the code Claude wrote AFTER you answered its questions and approved its plan.

```{r}
# PASTE CODE FROM CLAUDE HERE AND RUN IT


```

CHECK: the patient counts in your table should total 6,481, and patients with no patient type should still be there as a group.

## ALGORITHM DESIGN PART 2

Compare the output from above with the output from this structured prompt.

- Task: Write R code to answer two questions. Q1. Does turnaround time differ by patient type? Q2. Are we meeting a 3-hour collect-to-verify goal?

- Context: I am a core lab director answering a concern from an ICU physician, who says ICU chemistry results come back slower than results for other units. Our target is \_\_\_\_% of patients within 3 hours. I am new to R.

- Material: My data is in a file called chem_subset.csv. I can't upload it. The columns are: \[paste the column names\]. One row is ***. Each patient has*** \_ rows. The columns col_rec_tat and rec_ver_tat are turnaround times in ***.***

- ***Constraints: Use dplyr. Count patients, not rows. For patients with no patient type,*** . Tell me anything you assume.

- Output: First give me your plan as numbered steps. Do not write code until I approve the plan. Then give me one table by patient type with the number of patients, the median total turnaround time and the percent within 3 hours, plus one sentence answering each question. Keep the code short, with a comment on each step.

- If anything essential is missing, ask me before answering.

Use the code chunk below to evaluate all or part of the output.

```{r}

```

### DEBRIEF. Does this make sense?

```{r}
n_distinct(chem_subset$mrn) # how many patients are in the data? Do they total 6,481?

nrow(chem_subset)/n_distinct(chem_subset$mrn)    # how many rows per patient? Is there 3 (one for each test)


```

If the total number of patients is 5,393, patients with no patient type were dropped.
